An elaborate data set on human gait and the effect of mechanical perturbations

### 2014-12-14 paper.tex
Proofed raw data section.

\section*{Raw Data}
%
The raw data consists of a set of ASCII tab delimited text files output from
both the ``mocap'' and ``record'' modules in D-Flow in addition to a YAML file
that contains all of the necessary meta data for the given trial. These three
files are stored in a hierarchy of directories with one trial per directory.
The directories are named in the following fashion \verb+T001/+ where \verb+T+
stands for ``trial'' and the following three digits are provide a unique trial
identification number.

The text file output from the mocap module in D-Flow is a tab delimited file.
The first line is the header and contains a time stamp column, frame number
column, marker position columns, force plate force/moment columns, force plate
center of pressure columns, other analog columns, and potentially results from
the real time Human Body Model \cite{Bogert2013} which is included with the
D-Flow software. These are detailed below. The numerical values of the
measurements are provided in decimal floating point notation with 6 decimals of
precision, e.g. 123456.123456 [\%1.6f].

@@

The output from the D-Flow mocap module is stored in a tab delimited file named
\verb+mocap-xxx.txt+ where \verb+xxx+ represents the trial id number. The file
contains several columns:

\begin{description}
  \item[TimeStamp] The monotonically increasing computer clock time when D-Flow
    receives a frame from Cortex. These are approximately at 100 hz and given
    in seconds.
  \item[FrameNumber] Monotonically increasing positive integers that correspond
    to each frame received from Cortex.
  \item[Marker Coordinates] Any column that ends in \verb+.PosX+, \verb+.PosY+,
    or \verb+.PosZ+ are marker coordinates expressed in Cortex's Cartesian
    reference frame. These values are in meters.
  \item[Ground Reaction Loads] There are three ground reaction forces and three
    ground reaction moments recorded by each of the two force plates in Newtons
    and Newton-Meters, respectively. The prefix for these columns is either
    \verb+FP1+ or \verb+FP2+ and represents either force plate 1 (left) or 2
    (right). The suffixes are either \verb+.For[XYZ]+, \verb+.Mom[XYZ]+ for the
    forces and moments, respectively. The force plate voltages are sampled at a
    much higher frequency than the cameras, but delivered at the Cortex camera
    sample rate, 100 Hz through the D-Flow mocap module. A force/moment
    calibration matrix stored in Cortex converts the voltages to forces and
    moments before sending it to D-Flow [1]. Cortex also computes the center of
    pressure from the forces, moments, and force plate dimensions. These have
    the same prefixes for the plate number, have the suffix \verb+.Cop[XYZ]+,
    and are in meters.
  \item[Analog Channels] Several analog signals are recorded under column
    headers \verb+Channel[1-99].Anlg+. These correspond to analog signals
    sampled by Cortex and correspond to the 96 analog-channel National
    Instruments 6255 DAQ box. The first twelve are the voltages from the
    force plate load cells. We also record the acceleration of 4 points on the
    treadmill base in channels 61-72.  TvdB: are these useful?  If so, we need to
    provide information on how to process these channels. If not, maybe
    not mention them at all?
\end{description}

\subsection*{record-xxx.txt}
%
The record module also outputs a tab delimited ASCII text file. It includes a
\verb+Time+ column which records the D-Flow system time in seconds which
corresponds to the time recorded in the \verb+TimeStamp+ column in mocap module
tsv file. There are two additional columns \verb+RightBeltSpeed+ and
\verb+LeftBeltSpeed+ which provide the independent belt speeds measured in
meters per second by an encoder in the treadmill.

Finally, the record module is capable of recording the time at which various
events occur, as detected or set by D-Flow. It does this by inserting commented
(\#) lines in between the rows when the event occurred. The record files have
several events that correspond to the phases of the protocol:

\begin{description}
  \item[A: Force Plate Zeroing] Marks the time at which there is no load on the
    force plates and when the voltages were zeroed.
  \item[B: Calibration Pose] Marks the time at which the person is in the
    T-Pose.
  \item[C: First Normal Walking] Marks the time when the treadmill begins Phase
    1: constant belt speed.
  \item[D: Longitudinal Perturbation] Marks the time when the treadmill begins
    Phase 2: longitudinal perturbations in the belt speed.
  \item[E: Second Normal Walking] Marks the time when phase 3 starts: constant
    belt speed.
  \item[F: Unloaded End] Marks the time at which there is no load on the force
    plates and the belts are stationary.
\end{description}

@@

The \verb+study+ section contains identifying information for the overall
study, an identification number, name, and description. This is the same for
all meta data files in the study. Details are given below:

\begin{description}
  \item[id] An integer specifying a unique identification number of the study.
  \item[name] A string giving the name of the study.
  \item[description] A string with a basic description of the study.
\end{description}

The \verb+subject+ provides key value pairs of information about the subject in
that trial. Each subject has a unique identification number along with basic
anthropomorphic data. The following details the possible meta data for the
subject:

\begin{description}
  \item[age] An integer age in years of the subject at the time of the trial.
  \item[ankle-width-left] A float specifying the width of the subjects left
    ankle.
  \item[ankle-width-right] A float specifying the width of the subjects right
    ankle.
  \item[ankle-width-units] A string giving the units of measurement of the
    ankle widths.
  \item[id] An unique identification integer for the subject.
  \item[gender] A string specifying the gender of the subject.
  \item[height] A float specifying the measured height of the subject (with
    shoes on) at the time of the trial.
  \item[height-units] A string giving the units of the height measurement.
  \item[knee-width-left] A float specifying the width of the subjects left
    knee.
  \item[knee-width-right] A float specifying the width of the subjects right
    knee.
  \item[knee-width-units] A string giving the units of measurement of the
    knee widths.
  \item[mass] A float specifying the self-reported mass of the subject.
  \item[mass-units] A string specifying the units of the mass measurement.
\end{description}

The \verb+trial+ section contains the information about the particular trial.
Each trial has a unique identification number along with a variety of other
information, detailed below:

\begin{description}
  \item[analog-channel-map] A mapping of the strings D-Flow assigns to signals
    emitted from the analog channels of the NI USB-6255 to names the user
    desires.
  \item[cortex-version] The version of Cortex used to record the trial.
  \item[datetime]A date formatted string giving the date of the trial in the
    \verb|YYYY-MM-DD| format.
  \item[dflow-version] The version of D-Flow used to record the trial.
  \item[events] A key value map which prescribes names to the alphabetic events
    recorded in the record file.
  \item[files] A key value mapping of files associated with this trial
    where the key is the D-Flow file type and the value is the path to the file
    relative to the meta file.
  \item[hardware-settings] There are tons of settings for the hardware in both
    D-Flow, Cortex, and the other software in the system. This contains any
    non-default settings.
    \begin{description}
      \item[high-performance] A boolean value indicating whether the D-Flow high
        performance setting was on (True) or off (False).
    \end{description}
  \item[id] An unique three digit integer identifier for the trial. All of the
    file names and directories associated with this trial include this number.
  \item[marker-map] A key value map which maps marker names in the mocap file
    to the user's desired names for the markers.
  \item[marker-set] Indicates the HBM~\cite{Bogert2013} marker set used during
    the trial, either full, lower, or NA.
  \item[nominal-speed] A float representing the nominal desired treadmill speed
    during the trial.
  \item[nominal-speed-units] (string) The units of the nominal speed.
  \item[notes] Any notes about the trial.
  \item[pitch] A boolean that indicates if the treadmill pitch degree of
    freedom was actuated during the trial.
  \item[stationary-platform] A boolean that indicates whether the treadmill
      sway or pitch motion was actuated during the trial. If this flag is
      false, the measured ground reaction forces must be compensated for the
      inertial affects and be expressed in the motion capture reference frame.
  \item[subject-id] An integer corresponding to the subject in the trial.
  \item[sway] A boolean that indicates if the treadmill lateral (sway) degree
    of freedom was actuated during the trial.
\end{description}

\begin{listing*}
  \small
  \begin{minted}[gobble=4]{yaml}
    study:
       id: 1
       name: Gait Control Identification
       description: Perturb the subject during walking and running.
    subject:
       id: 8
       age: 20
       mass: 70.0
       mass-units: kilograms
       height: 1.572
       height-units: meters
       knee-width-left: 107.43
       knee-width-right: 107.41
       knee-width-units: millimeters
       ankle-width-left: 70.52
       ankle-width-right: 67.66
       ankle-width-units: millimeters
       gender: male
    trial:
       id: 58
       subject-id: 8
       datetime: 2014-03-28
       notes: >
           The subject did a somersault during this trial instead of following
           instructions to walk. Will have to use for another study.
       nominal-speed: 0.8
       nominal-speed-units: meters per second
       stationary-platform: True
       pitch: False
       sway: False
       hardware-settings:
           high-performance: True
       dflow-version: 3.16.1
       cortex-version: 3.1.1.1290
       marker-map:
           M1: LHEAD
           M2: THEAD
           M3: RHEAD
           M4: FHEAD
           M5: C7
       analog-channel-map:
           Channel1.Anlg: F1Y1
           Channel2.Anlg: F1Y2
           Channel3.Anlg: F1Y3
           Channel4.Anlg: F1X1
       events:
           A: Force Plate Zeroing
           B: Calibration Pose
           C: First Normal Walking
           D: Longitudinal Perturbation
           E: Second Normal Walking
           F: Unloaded End
       files:
           compensation: ../T057/mocap-057.txt
           mocap: mocap-058.txt
           record: record-058.txt
           meta: meta-058.yml
  \end{minted}
  \caption{A fictitious example of a YAML formatted meta data file. All of the
    possible keys in the data set are shown.}
  \label{lis:example-meta-data}
\end{listing*}

>>>>

\section*{Raw Data}
%
The raw data consists of a set of ASCII tab delimited text files output from
both the ``mocap'' and ``record'' modules in D-Flow in addition to a manually
generated YAML file that contains all of the necessary meta data for the given
trial. These three files are stored in a hierarchy of directories with one
trial per directory. The directories are named in the following fashion
\verb+T001/+ where \verb+T+ stands for ``trial'' and the following three digits
are provide a unique trial identification number.

@@

The output from the D-Flow mocap module is stored in a tab delimited file named
\verb+mocap-xxx.txt+ where \verb+xxx+ represents the trial id number. The file
is tab delimited and contains a number of time series. The numerical values of
the time series are provided in decimal fixed point notation with 6 decimals of
precision, e.g. \verb|123456.123456|, regardless of the units. The first line
of the file holds the header. The header includes time stamp column, frame
number column, marker position columns, force plate force/moment columns, force
plate center of pressure columns, other analog columns, and potentially results
from the real time Human Body Model \cite{Bogert2013} which is included with
D-Flow. The columns are further described below:
%
\begin{description}
  \item[TimeStamp] The monotonically increasing computer clock time when D-Flow
    receives a frame from Cortex. These are recorded at approximately at
    100~\si{\hertz} and given in seconds.
  \item[FrameNumber] Monotonically increasing positive integers that correspond
    to each frame received from Cortex.
  \item[Marker Coordinates] Any column that ends in \verb+.PosX+, \verb+.PosY+,
    or \verb+.PosZ+ are marker coordinates expressed in Cortex's Cartesian
    reference frame. The prefixes match the marker labels given in
    Table~\ref{tab:marker-labels}. These values are in meters.
  \item[Ground Reaction Loads] There are three ground reaction forces and three
    ground reaction moments recorded by each of the two force plates in Newtons
    and Newton-Meters, respectively. The prefix for these columns is either
    \verb+FP1+ or \verb+FP2+ and represents either force plate 1 (left) or 2
    (right). The suffixes are either \verb+.For[XYZ]+, \verb+.Mom[XYZ]+ for the
    forces and moments, respectively. The force plate voltages are sampled at a
    much higher frequency than the cameras, but delivered at the Cortex camera
    sample rate, \~100~\si{\hertz} through the D-Flow mocap module. A
    force/moment calibration matrix stored in Cortex converts the voltages to
    forces and moments before sending it to D-Flow. Cortex also computes
    the center of pressure from the forces, moments, and force plate
    dimensions. These have the same prefixes for the plate number, have the
    suffix \verb+.Cop[XYZ]+, and are given in meters.
  \item[Analog Channels] Several analog signals are recorded under column
    headers \verb+Channel[1-99].Anlg+. These correspond to analog signals
    sampled by Cortex and correspond to the 96 analog channels in the National
    Instruments USB-6255. The first twelve are the voltages from the
    force plate load cells. We also record the acceleration of 4 points on the
    treadmill base in analog channels 61-72 that were in place in case inertial
    compensation for the lateral treadmill movement was required.
\end{description}

\subsection*{record-xxx.txt}
%
The record module also outputs a tab delimited ASCII text file with numerical
values at six decimal digits. It includes a \verb+Time+ column which records
the D-Flow system time in seconds. This time corresponds to the time recorded
in the \verb+TimeStamp+ column in mocap module tsv file which is necessary for
time synchronization. There are two additional columns \verb+RightBeltSpeed+
and \verb+LeftBeltSpeed+ which provide the independent belt speeds measured in
meters per second by a factory installed encoder in the treadmill.

Additionally, the record module is capable of recording the time at which
various preprogrammed events occur, as detected or set by D-Flow. It does this
by inserting commented (\verb|#|) lines in between the rows when the event
occurred. The record files have several events that delineate the different
phases of the protocol:
%
\begin{description}
  \item[A: Force Plate Zeroing] Marks the time at the beginning of the trial at
    which there is no load on the force plates and when the force plate
    voltages were zeroed.
  \item[B: Calibration Pose] Marks the time at which the person is in the
    calibration pose.
  \item[C: First Normal Walking] Marks the time when the treadmill begins Phase
    1: constant belt speed.
  \item[D: Longitudinal Perturbation] Marks the time when the treadmill begins
    Phase 2: longitudinal perturbations in the belt speed.
  \item[E: Second Normal Walking] Marks the time when phase 3 starts: constant
    belt speed.
  \item[F: Unloaded End] Marks the time at which there is no load on the force
    plates and the belts are stationary.
\end{description}

@@

The \verb+study+ section contains identifying information for the overall
study, an identification number, name, and description. This is the same for
all meta data files in the study. Details are given below:
%
\begin{description}
  \item[id] An integer specifying a unique identification number of the study.
  \item[name] A string giving the name of the study.
  \item[description] A string with a basic description of the study.
\end{description}

The \verb+subject+ section provides key value pairs of information about the
subject in that trial. Each subject has a unique identification number along
with basic anthropomorphic data. The following details the possible meta data
for the subject:
%
\begin{description}
  \item[age] An integer age in years of the subject at the time of the trial.
  \item[ankle-width-left] A float specifying the width of the subjects left
    ankle.
  \item[ankle-width-right] A float specifying the width of the subjects right
    ankle.
  \item[ankle-width-units] A string giving the units of measurement of the
    ankle widths.
  \item[id] An unique identification integer for the subject.
  \item[gender] A string specifying the gender of the subject.
  \item[height] A float specifying the measured height of the subject (with
    shoes and hat on) at the time of the trial.
  \item[height-units] A string giving the units of the height measurement.
  \item[knee-width-left] A float specifying the width of the subjects left
    knee.
  \item[knee-width-right] A float specifying the width of the subjects right
    knee.
  \item[knee-width-units] A string giving the units of measurement of the
    knee widths.
  \item[mass] A float specifying the self-reported mass of the subject.
  \item[mass-units] A string specifying the units of the mass measurement.
\end{description}

The \verb+trial+ section contains the information about the particular trial.
Each trial has a unique identification number along with a variety of other
information, detailed below:
%
\begin{description}
  \item[analog-channel-map] A mapping of the strings D-Flow assigns to signals
    emitted from the analog channels of the NI USB-6255 to names the user
    desires.
  \item[cortex-version] The version of Cortex used to record the trial.
  \item[datetime] A date formatted string giving the date of the trial in the
    \verb|YYYY-MM-DD| format.
  \item[dflow-version] The version of D-Flow used to record the trial.
  \item[events] A key value map which prescribes names to the alphabetic events
    recorded in the record file.
  \item[files] A key value mapping of files associated with this trial
    where the key is the D-Flow file type and the value is the path to the file
    relative to the meta file. The compensation file corresponds to an unloaded
    trial collected on the same day that could be used for inertial
    compensation purposes, if needed.
  \item[hardware-settings] There are tons of settings for the hardware in both
    D-Flow, Cortex, and the other software in the system. This contains any
    non-default settings.
    \begin{description}
      \item[high-performance] A boolean value indicating whether the D-Flow high
        performance setting was on (True) or off (False).
    \end{description}
  \item[id] An unique three digit integer identifier for the trial. All of the
    file names and directories associated with this trial include this number.
  \item[marker-map] A key value map which maps marker names in the mocap file
    to the user's desired names for the markers.
  \item[marker-set] Indicates the HBM~\cite{Bogert2013} marker set used during
    the trial, either full, lower, or NA.
  \item[nominal-speed] A float representing the nominal desired treadmill speed
    during the trial.
  \item[nominal-speed-units] A string providing the units of the nominal speed.
  \item[notes] Any notes about the trial.
  \item[pitch] A boolean that indicates if the treadmill pitch degree of
    freedom was actuated during the trial.
  \item[stationary-platform] A boolean that indicates whether the treadmill
      sway or pitch motion was actuated during the trial. If this flag is
      false, the measured ground reaction loads must be compensated for the
      inertial affects and be expressed in the motion capture reference frame.
  \item[subject-id] An integer corresponding to the subject in the trial.
  \item[sway] A boolean that indicates if the treadmill lateral degree of
    freedom was actuated during the trial.
\end{description}

\begin{listing}
  \small
  \begin{minted}[gobble=4]{yaml}
    study:
       id: 1
       name: Gait Control Identification
       description: Perturb the subject during walking and running.
    subject:
       id: 8
       age: 20
       mass: 70.0
       mass-units: kilograms
       height: 1.572
       height-units: meters
       knee-width-left: 107.43
       knee-width-right: 107.41
       knee-width-units: millimeters
       ankle-width-left: 70.52
       ankle-width-right: 67.66
       ankle-width-units: millimeters
       gender: male
    trial:
       id: 58
       subject-id: 8
       datetime: 2014-03-28
       notes: >
           The subject did a somersault during this trial instead of following
           instructions to walk. Will have to use for another study.
       nominal-speed: 0.8
       nominal-speed-units: meters per second
       stationary-platform: True
       pitch: False
       sway: False
       hardware-settings:
           high-performance: True
       dflow-version: 3.16.1
       cortex-version: 3.1.1.1290
       marker-map:
           M1: LHEAD
           M2: THEAD
           M3: RHEAD
           M4: FHEAD
           M5: C7
       analog-channel-map:
           Channel1.Anlg: F1Y1
           Channel2.Anlg: F1Y2
           Channel3.Anlg: F1Y3
           Channel4.Anlg: F1X1
       events:
           A: Force Plate Zeroing
           B: Calibration Pose
           C: First Normal Walking
           D: Longitudinal Perturbation
           E: Second Normal Walking
           F: Unloaded End
       files:
           compensation: ../T057/mocap-057.txt
           mocap: mocap-058.txt
           record: record-058.txt
           meta: meta-058.yml
  \end{minted}
  \caption{A fictitious example of a YAML formatted meta data file. All of the
    possible keys in the data set are shown.}
  \label{lis:example-meta-data}
\end{listing}
